Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2018-04-24 LatticeOptics/latticeOpticsV4.tex
Updating text, small edits here and there, repositioning figures, adding fig captions, replacing new figures

\author{A.M., S.B., F.H., M.F.P., C.S.C., G.J., D.G.}
\author{M. G.}
\affiliation{
  Department of Physics, Harvard University,
  Cambridge, Massachusetts, 02138, USA}

@@

Optical lattices holding an array of ultracold atoms have become a powerful experimental platform for a wide variety of investigations into various forms of quantum simulation \cite{Friedenauer2008, Kim2010, Struck2011, Simon2011, Yan2013, Drewes2016, Murmann2015}. %, strongly correlated systems, quantum state engineering, quantum computing and precision metrology.
Further, the recent advent of quantum gas microscopy~\cite{Bakr2009, Sherson2010, Haller2015, Cheuk2015, Parsons2015, Edge2015, Omran2015, Greif2016, Cheuk2016, Parsons2016, Boll2016, Cheuk2016a, Brown2016} has enabled studies that exhibit unprecedented control over few atom systems at extremely low energy scales, particularly those governed by the Hubbard Hamiltonian ~\cite{Hubbard1963}.
However, this progress has placed ever more stringent technical constraints on the lasers used to trap and manipulate the atomic systems.
For example, heating due to the intensity noise cannot be allowed to approach the longest thermalization timescales, lest the system fail to thermalize~\cite{Savard1997}.
Even if the system thermalizes, heating rates can limit the interrogation time available to probe a low-temperature state.
In a similar vein, experiments that rely on a small number of atoms held by multiple traps place stringent requirements on the relative positional stability of the traps, to ensure that the operations performed by the traps have adequate fidelity.%~\cite{Mazurenko2016}.FIXME: CITATIONS

To address these concerns, we discuss a high power optical lattice system that exhibits levels of intensity noise below $-120\,$dBc between $1\,$kHz and $3\,$MHz.
The stability of the lattice position is below the single site level and the stability of the harmonic confinement is at the few site level.
\begin{figure}
  \begin{center}
    \includegraphics{fig/figure1/figure1.pdf}
    \caption{\textbf{High level overview} showing the major components of the lattice laser feedback system.}\label{fig:high_level}
  \end{center}
\end{figure}

@@

@@

@@

\begin{figure*}
  \begin{center}
    \includegraphics{fig/optical_layout_wide.pdf}
    \caption{\textbf{Optical layout of the optical lattice system for beam shaping and modulation.}}\label{fig:optical_layout}
  \end{center}
\end{figure*}

@@

\begin{figure*}
  \begin{center}
    \includegraphics{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop}. This Berek compensator based feedback loop is used to vary the beam intensity at relatively high power. (a) Shows the small signal impulse response and (b-c) show the corresponding Bode plots.}\label{fig:berek_step_response}
  \end{center}
\end{figure*}
A key point, as seen in figure \ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator. 
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in figure \ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate. 
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.
Since two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously. 
To accomplish this, the low power loop is always manually railed to its high extreme when the high power loop is active, and the digital high power loop is disabled when the low power loop is active.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{fig/averaged_mott.png}
    \caption{\textbf{Averaged images of Mott insulators}. A series of Mott insulators was created over the course of approximately one hour, and the average image is shown\cite{Greif2016}. If the lattice fluctuated across more than a small fraction of a lattice site, the modulation contrast would be washed out. The absence of this effect means that the phase of the optical lattice is stable.}\label{fig:averaged_mott}
  \end{center}
\end{figure}

@@

\section{Long-term stability of the lattice structure}
FIXME: I PROPOSE MOVING THIS INTO THE SUPPLEMENTARY MATERIALS: IT MAY BE INTERESTING TO EXPERTS, BUT IS A LITTLE TOO ``ANECDOTAL'' FOR THE MAINTEXT.
A benefit of site-resolved imaging is that every measurement of the atomic distribution is also a direct measurement of the underlying lattice structure. 
As discussed in the supplementary material of~\cite{Parsons2016, Greif2016}, the Fourier transform of an atomic distribution contains distinct peaks corresponding to the lattice structure, and these peaks can be used to extract the phase, relative angle, and spacing of the lattices.
For this reason, we looked back at nine months of data spanning June 2016 through April 2017 in order to learn how these parameters change over time.
Several interesting features can be seen from this figure \ref{fig:long_term_stability}.
\begin{enumerate}
  \item During the summer of 2016, the lattice spacing appears to have slowly varied over as much as 10\%.
    This can be attributed to unexplained drifts of the optical system that required daily refocusing of the imaging system using an auxiliary lens, which simultaneously changed the magnification. 
  \item The imaging system was simplified in September 2016, causing a sudden change in the magnification. 
  \item The system has been more stable in the spring of 2017 - the lattice spacing $|v|$ exhibits fractional fluctuations $\sigma_{|v|}/\mu_{|v|}\leq 0.005$.
  \item Unlike magnification, which sensitively depends on the imaging system focus, the angle is unlikely to be affected by the imaging system unless the camera is mechanically rotated. We measure that in the spring of 2017, the relative angle of the lattices has exhibited fluctuations with an RMS of $0.06^\circ$.
\end{enumerate}

\begin{figure*}
  \begin{center}
    \includegraphics{fig/lattice_stability.pdf}
    \caption{\textbf{Long term positional stability over approximately nine months.}}\label{fig:long_term_stability}
  \end{center}
\end{figure*}

\section{Conclusion}
We have presented the design and specifications of an optical lattice system that has been successfully used for quantum gas microscopy of low temperature phases of the Hubbard model.
In this design, key emphasis has been placed on the stability of the pointing and the intensity of the optical lattice, of particular importance to the underlying physics.
To improve performance, the system can be improved in a variety of ways, such as by:
\begin{enumerate}
  \item Replacement of the \textit{Nufern} fiber amplifiers with a lower noise model.
  \item Replacement of the analog feedback loop by a low-noise digital loop.
  \item Active stabilization of the laser pointing.
\end{enumerate}
Even without these improvements, the system has exhibited robust and reliable performance in performing a wide variety of experiments.

>>>>

\author{A.Mazurenko, S. Blatt, F. Huber, M. F. Parsons, C. S. Chiu, G. Ji, D. Greif}
\author{M. Greiner}
\affiliation{
  Department of Physics, Harvard University,
  Cambridge, Massachusetts, 02138, USA}

@@

Optical lattices holding an array of ultracold atoms have become a powerful experimental platform for a wide variety of investigations into various forms of quantum simulation \cite{Friedenauer2008, Kim2010, Struck2011, Simon2011, Yan2013, Drewes2016, Murmann2015}. 
The recent advent of quantum gas microscopy~\cite{Bakr2009, Sherson2010, Haller2015, Cheuk2015, Parsons2015, Edge2015, Omran2015, Greif2016, Cheuk2016, Parsons2016, Boll2016, Cheuk2016a, Brown2016} has enabled studies that exhibit unprecedented control over individual atoms at extremely low energy scales, particularly those governed by the Hubbard Hamiltonian ~\cite{Hubbard1963}.
However, this progress has placed ever more stringent technical constraints on the lasers used to trap and manipulate the atomic systems.
For example, heating due to the intensity noise cannot be allowed to approach the longest thermalization timescales, lest the system fail to thermalize~\cite{Savard1997}.
Even if the system thermalizes, heating rates can limit the interrogation time available to probe a low-temperature state.
Similarly, experiments that rely on a small number of atoms held by multiple traps place stringent requirements on the relative positional stability of the traps, to ensure that the operations performed by the traps have adequate fidelity.

In this work we discuss a high power optical lattice system that exhibits levels of intensity noise below $-120\,$dBc between $1\,$kHz and $3\,$MHz.
The stability of the lattice position is below the single site level and the stability of the harmonic confinement is at the few site level.

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.85\columnwidth]{fig/figure1/figure1.pdf}
    \caption{\textbf{High level overview} showing the major components of the lattice laser feedback system.}\label{fig:high_level}
  \end{center}
\end{figure}

@@

In the following we give an in-depth technical description of the optical lattice setup we developed. We start with the laser system itself which includes the seed laser, amplification and optical setup for controlling the laser beam. In the next section we provide a detailed description of the two-stage laser power feedback system. In the last two sections we conduct a performance analysis of the short- and long-term stability of the optical lattice.

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics{fig/optical_layout_wide.pdf}
    \caption{\textbf{Optical layout of the optical lattice system for beam shaping and modulation.}}\label{fig:optical_layout}
  \end{center}
\end{figure*}

@@

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop}. This Berek compensator based feedback loop is used to vary the beam intensity at relatively high power. (a) Shows the small signal impulse response and (b-c) show the corresponding Bode plots.}\label{fig:berek_step_response}
  \end{center}
\end{figure}
A key point, as seen in figure \ref{fig:berek}, is that if the waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator. 
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a digital feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in figure \ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate. 
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.
Since two feedback loops are present in the system (the second one detailed in the next section), care must be taken so that they do not function simultaneously. 
To accomplish this, the low power loop is always manually railed to its high extreme when the high power loop is active, and the digital high power loop is disabled when the low power loop is active.

@@

\begin{figure*}
  \begin{center}
    \includegraphics{fig/AveragedMI.pdf}
    \caption{\textbf{Averaged images of Mott insulators}. a) Several Mott insulators were created over the course of approximately one hour, and the average image is shown\cite{Greif2016}. If the lattice fluctuated across more than a small fraction of a lattice site, the modulation contrast would be washed out. The absence of this effect means that the phase of the optical lattice is stable. b) shows an individual image of a Mott insulator.}\label{fig:averaged_mott}
  \end{center}
\end{figure*}

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\columnwidth]{fig/lattice_stability.pdf}
    \caption{\textbf{Long term stability over approximately nine months.} The lattice spacing $|\mathrm{v}_x|$ and $|\mathrm{v}_y|$ and the lattice angles $\theta_x$ and $\theta_y$ along the two lattice axes $x$ and $y$ are shown over the duration of almost 1 year. }\label{fig:long_term_stability}
  \end{center}
\end{figure}

\section{Long-term stability of the lattice structure}
A benefit of site-resolved imaging is that every measurement of the atomic distribution is also a direct measurement of the underlying lattice structure. 
As discussed in the supplementary material of~\cite{Parsons2016, Greif2016}, the Fourier transform of an atomic distribution contains distinct peaks corresponding to the lattice structure, and these peaks can be used to extract the phase, relative angle, and spacing of the lattices.
For this reason, we recorded a long-term measurement spanning almost one year in order to learn how these parameters change over time, see Fig.~\ref{fig:long_term_stability}.
Several interesting features can be seen from the observed behavior.
\begin{enumerate}
  \item During the first few months the lattice spacing appears to have slowly varied over as much as $10\%$.
    This might be caused by drifts of the optical system that required daily refocusing of the imaging system using an auxiliary lens, which simultaneously changed the magnification. 
  \item The imaging system was simplified in September 2016, causing a sudden change in the magnification. 
  \item The system has been more stable since then and the lattice spacing $|v|$ exhibits fractional fluctuations $\sigma_{|v|}/\mu_{|v|}\leq 0.005$.
  \item Unlike magnification, which sensitively depends on the imaging system focus, the angle is unlikely to be affected by the imaging system unless the camera is mechanically rotated. We measure that in the spring of 2017, the relative angle of the lattices has exhibited fluctuations with an RMS of $0.06^\circ$.
\end{enumerate}

\section{Conclusion}
We have presented the design and specifications of an optical lattice system that has been successfully used for quantum gas microscopy of low temperature phases of the Hubbard model.
In this design, key emphasis has been placed on the stability of the pointing and the intensity of the optical lattice, of particular importance to the underlying physics.
To improve performance, the system can be upgraded in a variety of ways. For example, the \textit{Nufern} fiber amplifiers can be replaced with a lower noise model. Or the analog feedback loop can be replaced by a low-noise digital version. Additionally, active stabilization of the laser pointing could be implemented in the future.
